A new approach to bias correction in high-throughput sequencing data

### 2011-09-30 response/response.tex

\documentclass{article}
\usepackage{amsmath}
\usepackage{dcj}

\begin{document}

We appreciate the time and effort extended by the anonymous reviewers and the
editorial staff to evaluate the original version of this manuscript (titled
``Correcting for bias in high-throughput sequencing data''), which was submitted
on July 8. Upon review, the second of the two reviewers raised some concerns
leading to its rejection. However, we believe that the bulk of these concerns
are either incongruous with the content of the manuscript or addressable with
very little effort, and therefore do not warrant its rejection.  Here we have
provided a detailed response to these comments, and have included additional
material in an effort to clarify these points.

The second reviewer raised three major concerns. We summarize our
responses to these issues here.  The first concern raised is
that the method we present is unjustifiably complex, involving a ``plethora of
tuning parameters'', and that the much simpler method of Hansen, et al., would
perform as well if trained on enough data.  In the data we presented, the simple
method developed by Hansen, et al., performed poorly regardless of the number of
reads used in training, which varied from hundreds of thousands to tens of
millions.  The methods presented by Li, et al., are more effective, yet involve
more parameters, and require more time to train, so that our method, while often
being more effective, also offers a significant simplification.
Additionally, we believe the structure of our models give some
theoretical insight into these differences in efficacy.

@@

Secondly, it is asserted that we did not demonstrate that uniformity of read
coverage is increased after correcting for bias using our method.  Yet, in
section 3.2 we had presented a rigorous statistical analysis over a large and
diverse collection of data sets showing that this is indeed the case. This
section of the paper was perhaps overlooked by the reviewer, as it received
no mention in this criticism.

Finally, the reviewer doubts the biological relevance of the bias correction
methods. Having demonstrated an increase in uniformity (section 3.2) and a
decrease in nucleotide frequency divergence (section 3.1), we believe a
compelling case was made that correcting for bias increases the accuracy of
quantification. In this revision we have supplied another line of evidence:
analysis showing that correlation between RNA-Seq and qRT-PCR is increased after
correcting for bias.

Under these circumstances we respectfully request that the manuscript, in this
revised form, be reconsidered for publication. Detailed responses to the
reviewers concerns are presented below.

@@

\subsection*{Response to Reviewer 1}

\begin{quote}
(1) Page 3, 2.2.1 mentions a correction for bias introduced by PCR amplification.
RNA sequencing methods are available where PCR amplification is not used.
Correcting for bias in these cases removes true duplication.  Could the authors
please comment on this?
\end{quote}

First, we emphasize that PCR amplification is only one of a number of technical steps
involved in sequencing that potentially introduce bias.  Our method is agnostic
to the source of bias, and because model complexity is penalized, a very sparse,
or even empty model will be trained on a data set without bias.  Thus, if PCR
amplification is responsible for a majority of the bias, and an RNA-Seq method
without PCR amplification is used producing largely bias-free data, then using
the method to correct for bias will have little if any effect.

@@

As a concrete example, we have included new data examining the FRT-Seq protocol
of Mamanova, et al., in which no PCR-amplification is performed during library
preparation. We observe that bias is still present in this data set, albeit at
lower levels than other uncorrected RNAseq data examined.  Our method accounts
for this reduced bias by constructing a much sparser model, relative to the
other RNA-Seq data sets we have examined.

We have included additional material in Supplementary Section 10 exploring this
point in detail. We first prove an upper bound on the probability of the model
predicting the presence of bias when there is none, demonstrating that is less
than 0.0004 when more than 10,000 reads are used. We then investigate this
empirically by simulating random reads, drawn uniformly from exonic sequences
within the human genome.  We trained our model 100 times with 10,000 simulated
reads, and in no instance was bias predicted (i.e., a non-empty model trained).
We repeated this using only 100 reads, and in only 3 out of 100 trials was
bias incorrectly predicted.

@@

It should also be noted that, while the brute-force method of Hansen, et. al.,
has fewer tuning parameters, the other methods evaluated in the manuscript (the
GLM and MART methods of Li, et. al.), involve more parameters than the method we
have described. (For example, training the MART method involves parameters
controlling the complexity penalty, the number of trees, the lower bound
threshold for read counts, in addition to others that are not described or
easily changed from the values divined by the authors.  Additionally, their
method requires selection of a set of annotated genes for training.)  Our method
therefore presents a simplification over competing methods.

\begin{quote}
To justify this cost, the
authors make the argument that their approach avoids the potential
overfitting of a brute force approach such as Hansen's. However, I am
not convinced: they evaluate both methods, theirs and Hansen's, on
'more than 600,000 reads', yet this seems unrealistic: current
datasets have hundreds of millions of reads, and that may just solve
the estimation precision problem in the simple brute force approach.

(I appreciate that the implementation of the brute force approach in
the Genominator package is ridiculously inefficient, but that can
easily be fixed by more competent programming.)
\end{quote}

We regret that we were not more accurate or explicit on this point. The 600,000
figure somewhat underestimated the actual number used. Table 2 now given the
correct numbers of reads used to train each method. The minimum number of reads
used to train the Hansen method was in fact 1.9 million, and we used as many as
23 million.  (As noted in the paper, we did reimplement the Genominator method
in order to do this.)

@@

We do not wish to ``pick on'' Hansen, et al.  Theirs was a pioneering
paper and presents a simple, intuitive approach for tackling the
problem they identified.  Nevertheless, their brute-force method
performed poorly on every data set we tested, even the largest ones,
suggesting that simply throwing more data at a suboptimal solution
will not adequately solve the problem.  In our opinion, the results we
presented were unambiguous: the other methods perform far better with
far fewer reads.

%Moreover, even if the brute-force method were effective if trained
%with many hundreds of millions of reads, many (possibly most) existing
%data sets are not this large, rendering it generally impractical.

Understanding this failure of a simple, appealing method is important for making
progress, and we feel that our paper provides a theoretical basis for
understanding these differences in efficacy, and we have amplified these points
in the revision.  Intuitively, one expects model accuracy to increase as the
number $n$ of training reads increases, but to decrease as the number $r$ of
model parameters grows, since the training data is ``spread more thinly.''  For
the particular case of estimating a multinomial distribution (as in Hansen),
Supplementary Section 9 quantifies this intuition very precisely, showing that
the expected KL divergence of the learned model with respect to the true
distribution very closely approximates $(r-1)/(2n)$, where $r=4^7=16384$ is the
number of model parameters learned.  The same analysis does not directly apply
to the MART, GLM or BN models, but we expect their behavior to be qualitatively
similar.  Thus, to the extent that a more parsimonious model such as ours
adequately captures the structure of the data, it is potentially orders of
magnitude more accurate for a given quantity of training data, simply by virtue
of having only a few hundred parameters learned from counts (Figure 3), as
opposed to tens of thousands.

Of course, training data is plentiful in NGS data sets, so, pragmatically, the
above analysis probably identifies a contributor but not the root cause of
Hansen et al.'s relatively poor performance in our experiments. Rather, we
believe the problem is that \emph{bias is not confined to the initial
heptamer}.The model structures illustrated in Figure 3 show that nucleotides
well outside that region are often biased.  (The the theoretical results and
randomization studies sketched above, and presented in Supplementary Section 10
show that our model building algorithm is not prone to spurious discovery of
overly-complex models.)  Furthermore, these examples show that the read sampling
is often biased by positions \emph{in front of the read}, which are not
currently considered by Hansen.  Their method could be modified easily to
consider $k$-mers spanning the read start, but the entropy analysis in
Supplementary Section 9 shows that extending it to $k \gg 7$ is infeasible,
since the exponential growth in $r=4^k$ will destroy the model's accuracy even
in large data sets. E.g., $k \ge 20$ would be necessary to cover the biased
regions in all data sets shown in Figure 3, necessitating trillions of training
reads to build an adequate model by Hansen's method.

@@

\begin{quote}
The major weakness of the ms, in my view, is the choice of 'figures of
merit' that the authors made (Fig.1, Sections 3.1, 3.2). What I would
have expected to see are

1.) Plots closer to the data such as Supplementary Fig.5 of
reference [1]. That figure shows highly uneven coverage of reads
within exons.
- Does this unevenness go away with the proposed adjustment?
\end{quote}

During preparation we chose not to include a figure such as the one suggested
by the reviewer as we believed it would not be legitimately informative. It
would be far too easy to cherry-pick a miniscule genomic interval that makes
even a very shoddy method appear favorable.

Instead, we presented data (in Section 3.2) that answers the reviewer's question
(i.e., ``does this unevenness go away with the proposed adjustment?'') in
rigorous and comprehensive manner. We implemented a Poisson regression metric,
which quantitatively measures ``uniformity'' by calculating the likelihood that
the reads sampled across a given exon were sampled from a uniform distribution.
Using this, we performed a large cross-validation study proving the
effectiveness of our method at increasing uniformity across thousands of exons
drawn from a diverse group of data sets (Figure 5), rather than qualitatively
across one exon.

In this revision, as suggested, we have revised Figure 2 to show an
example of this increase in uniformity.  We believe the figure is a
nice, graphic illustration of the benefits of our method and thank the
reviewer for causing us to reconsider it.  We stress, however, that
this figure represents a minute subset of the analysis that had
previously been presented, and is included only as an illustration.
The results in sections 3.1 and especially in 3.2 are far more
quantitative and comprehensive.

@@

\begin{quote}
- Also, as is evident from that figure, for some of the experimental
protocols, some regions within highly expressed exons get vanishingly
small read coverage.  These are the really hard situations, and much
more worrying than instances where coverage just fluctuates a bit.
How does the proposed scaling adjustment of Section 2.1 work in these
places (i.e. when $s_{i}$ is very small)?
\end{quote}

In Figure 2, we now indicate that the adjustments made by the method to read
counts tend to be small. In our test, 95\% of a sample of 10 million predicted
bias coefficients were between 0.07 and 4.27. This is also reflected indirectly
in the results of Section 3: extremely large adjustments would result in large
variance of the predictions, leading to poor performance in our evaluations,
which we do not see.

@@

\begin{quote}
2.) A 'bottom-line', biologically relevant figure of merit: E.g., does
the identification of differentially expressed genes improve? The
clustering of gene expression profiles? The identification of
transcript isoforms?
\end{quote}

In Section 3, we demonstrate, after correction, a significant improvement in
the two obvious anomalies associated with the issue of sequence bias:
non-uniform read coverage, and divergent nucleotide frequencies. We assess these
aspects in a statistically rigorous manner and with a large sample size. We
believe that this analysis presents a compelling case that bias correction is
worthwhile, and that the method we present is competitive with, and quite often
superior to existing methods.

In this revision, we have included more data (Section 3.3), showing, in addition
to the improvements in uniformity and nucleotide frequencies, significantly
increased correlation between RNA-Seq and qRT-PCR, after adjusting for bias in
RNA-Seq.  These results are consistent with, and reinforce our previous results.

@@

\begin{quote}
Minor points:
------------
1. The ms seems confused between classification and regression.
Reference [2] might be helpful to resolve some of that.
\end{quote}

The term ``classification'' was used in the context of noting that the algorithm
is adapted from one typically used in classification. It is correct to say that
the method here performs regression, and we have tried to clarify this language
in this revision.

\begin{quote}
2. The chosen Bayesian network is rather complicated, and a spectrum
of alternative methods exists between that and the brute force
approach of Hansen (say, lasso or svm regression). In absence of a
more meaningful benchmark, the choice of method seems arbitrary.
\end{quote}

We agree completely that a spectrum that a spectrum of methodhs are available.
Our motivation for adopting a Bayesian network model is the very natural
probabilistic interpretation that we have described in Section 2.1. We disagree
however with the reviewer's opinion that Bayesian networks are inherently more
complicated than Lasso regression or support vector machines.  In this
application, the latter two methods would generate output that is arguably less
transparent and interpretable, and would neither be significantly simpler to
implement, nor involve fewer parameters.  Our hypotheses (above) concerning the
reasons for the weakness of Hansen's method illustrate the benefit of this
approach; we do not believe svm regression, say, would have yielded similar
insights.

### 2011-09-30 seqbias-paper.tex
(almost) final draft
final draft

\title{Correcting for bias in high-throughput sequencing data}
\author[Jones \textit{et~al}]
{Daniel C. Jones\,$^{1,}$
\footnote{to whom correspondence should be addressed}\hspace{0.5em}
Walter L. Ruzzo\,$^{1,2,3}$
Xinxia Peng\,$^{4}$
Michael G. Katze$^{4}$
}

@@

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated by protocol-specific sequence bias. The sources of the bias are
unknown, but may influenced by PCR amplification, or differing primer affinities
and mixtures, for example.  The result is decreased accuracy in many
applications, such as de novo gene annotation and isoform quantification in
RNA-Seq, and peak calling in ChIP-Seq.

\section{Results:}
We present a method to measure and correct for these influences using a simple
graphical model. Our model does not rely on existing gene annotations, making it
applicable to any high throughput sequencing data. We evaluate our method on
several data sets, and by multiple criteria, demonstrating that it effectively
decreases bias and increases uniformity.

@@

In the last few years, RNA-Seq has emerged as a promising alternative to
microarrays in quantifying RNA abundance. But, as microarray technology has
brought with it technical challenges ranging from developing robust
normalization to accounting for cross-hybridization, RNA-Seq presents a new set
of problems.

A significant example is the effect of transcript length on quantification.  As
noted by \citet{Oshlack2009} transcript length biases discovery of
differential expression, since more reads are aligned to longer transcripts,
granting higher statistical significance in tests of differential expression.
Recently, some methods have been developed attempting to account for this effect
in gene set enrichment tests \citep{Young2010,Gao2011}, leading to more accurate
downstream analysis.

The underlying nucleotide sequence also influences quantification in a manner
that is complex and protocol-specific. In an ideal experiment, the number of
reads mapping to a particular position in the genome is a function of RNA
abundance and should not be additionally dependent on the sequence at that
position. As noted first by \citet{Dohm2008}, this is not the case.  As
illustration, Figure \ref{fig:freqs} plots this nonuniformity in nucleotide
frequencies on five data sets (see, Table \ref{tab:datasets}), each using a
different protocol.

@@

Another model, proposed by \citet{Hansen2010}, directly estimates the
distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution, sampled from the ends of reads. The read count
at a given position is then adjusted by the ratio of the foreground and
background heptamer probabilities. Specifying two distributions over heptamers
(i.e. foreground and background distributions) requires \mbox{$2(4^7-1) =
32,766$} parameters, so while no gene annotations are needed to train such a
model, a significant number of reads are required, and a number that increases
exponentially with $k$, if it were desirable to to model $k$-mers for $k > 7$.

@@

We begin with a natural model of an RNA-Seq experiment (and one that is often
assumed, whether implicitly or otherwise). The number of reads $x_i$ aligned to
genomic position $i$ is an unbiased estimate of RNA abundance. Furthermore, we
assume reads may be treated as independent and identically distributed samples.
That is, if $N$ reads are generated, and $m_i$ is the event that a generated
read maps to position $i$.
$$ E[x_i] = N \Pr[m_i] $$

@@

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias. The predicted bias can then be
used to adjust read counts, as in (e). In (d) and (e) we show the results of
this method applied to the $3'$ UTR of Apoa2, using data from
\citet{Mortazavi2008}. In bias coefficients predicted across 10 million positions
of chromosome 1, the log10 bias of 95\% of the positions were between -1.14 and
0.63, suggesting that most adjustments are not large.
}
\label{fig:overview}
\end{figure}

@@

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity heuristically using the Bayesian
Information Criterion \citep{Schwarz1978a}. Where $m$ is the number of
parameters needed to specify the model, we maximize, 
$$ \ell' = 2 \ell - m \log n $$

@@

The run time of the training procedure is further reduced in practice by
imposing the following two restrictions on the structure of the network,
\begin{enumerate}
\item The in-degree (i.e. number of parents) of any node must be less than some
number $p_{\text{max}}$.
\item $|j - i| \le d_{\text{max}}$ for all edges $(i,j)$ and some number $d_{\text{max}}$.
\end{enumerate}
The second of these encodes the assumption that distant nucleotides are
effectively independent. We choose $p_{\text{max}} = 4$ and $d_{\text{max}} =
10$, as reasonable default values.

@@

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing three assumptions we make of an ideal, unbiased experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}), measured
from reads within exons, should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across a single exon should follow, approximately, a Poisson
process.
\item Adjusting for bias should increase the agreement between RNA-Seq and
qRT-PCR.
\end{enumerate}

Evident from Figure \ref{fig:overview}, the assumption of uniform read coverage
often does not hold in typical RNA-Seq data sets. Although the bias corrected
read counts across the exon pictured in this example are visibly more uniform,
we sought a simple, objective tests that could be applied genome-wide. To this
end, we used cross-validation tests (i.e., methods were trained and tested on
disjoint subsets of the same RNA-Seq data sets) of a quantitative measure of the
increase in uniformity of nucleotide frequencies (Kullback-Leibler divergence in
Section 3.1) and increase in uniformity of read coverage (Poisson regression in
Section 3.2). Additionally, we compare RNA-seq based estimate of gene expression
to qRT-PCR based estimated for the same genes, showing increased correlation
between the two methods after bias correction (Section 3.3).

@@

\begin{table}
\processtable{Data sets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllcc}\toprule
 & & & Read & Num. of \\
Experiment & Species & Platform & length & reads \\\midrule
\citet{Wetterbom2010} & Chimpanzee & ABI & 33 & $3.9 \times 10^7$ \\
Katze (unpublished) & Macaque & ABI & 50  & $4.8 \times 10^7$ \\
\citet{Bullard2010} & Human & Illumina & 35 & $1.2 \times 10^7$ \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 & $3.9 \times 10^6$ \\
\citet{Trapnell2010} & Mouse & Illumina & 75 & $6.6 \times 10^7$ \\\botrule
\end{tabular}
}{}
\end{table}

@@

The differences in the methods being tested necessitated training procedures
unique to each.

@@

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as
$$D_k( f, f' ) = \sum_{x} \left( f_x \log_{2}( f_x / f'_x ) + f'_x \log_{2}( f'_x / f_x) \right)$$
where the sum is over all $k$-mers. This can be thought of a measure
dissimilarity between two probability distributions. If $f_x$ and $f'_x$ for a
$k$-mer $x$ are approximately equal, their log-ratio will also be approximately
zero, leading to a small KL divergence (zero, when the distributions are equal).
Conversely, very different $k$-mer frequencies will result in a larger KL
divergence.

@@

In this comparison, we measure how well the counts conform to a Poisson process.
The assumption of positional read counts following a Poisson distribution is
known to be a poor fit \citep{Srivastava2010}, but measuring the improvement in
the fit derived from correcting for bias remains a principled and easily
interpreted criterion. This increase in uniformity is illustrated in Figure
\ref{fig:overview}.

@@

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a prefect fit, occurring model being evaluated
achieves a likelihood of 1. Smaller number indicate an increasingly worse fit,
with $R^2 = 0$ representing no improvement over the null model, and $R^2 = 0.5$,
for example, indicating the model has a log-likelihood equal to half that of the
null model (a large improvement, as log-likelihoods negative or zero). This
measure can also take on values less than 0, indicating that the model has worse
fit than the null model (i.e., when adjusting read counts according the bias
coefficients leads to less uniform read coverage).

We compute $R^2$ for each of the test exons, giving us a sense of the variability
of the effectiveness of each model. The results of this analysis are plotted in
Figure \ref{fig:pois}.  To summarize each model with a single number, we can
examine the median $R^2$ value, as listed in Table \ref{tab:pois}.

\begin{figure}
\centerline{\includegraphics[width=0.30\textwidth]{pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we compute a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive, and increases as uniformity is
increased, and negative when uniformity is decreased.  Marked with asterisks are
methods over which the BN approach showed a statistically significant improvement
when applied to the same data, according to a one-sided Wilcoxon signed-rank
test. In each of those marked, we observed p-values less than $10^{-23}$.  Boxes
are plotted to mark the 25\%, 50\%, and 75\% quantiles, with whiskers extending
to 1.5 times the inter-quartile range (i.e., the span between the 25\% and 75\%
quantiles), and dots marking more extreme values.
}

    \label{fig:pois}
\end{figure}

\begin{table}
\processtable{
    The median $R^2$ goodness of fit statistic across test exons.
\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & \textbf{0.174} & 0.016 & 0.066 & -0.079 \\
Katze     & \textbf{0.280} & 0.243 & 0.158 & 0.033 \\
Bullard   & \textbf{0.267} & 0.163 & 0.224 & 0.157 \\
Mortazavi & \textbf{0.240} & 0.210 & 0.197 & 0.091 \\
Trapnell  & \textbf{0.289} & \textbf{0.289} & 0.248 & 0.138
    \end{tabular}
}
{
    The $R^2$ statistic measures increased uniformity in read coverage, after
    correcting for bias.  Here the median $R^2$ across the test exons is listed
    for each method and sample. A higher $R^2$ indicates a better fit.
}
\end{table}

@@

We used sequencing data previously published by \citet{Au2010} to evaluate the
effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
made available by the the Microarray Quality Control project \citep{Shi2006}. The
RNA-Seq data shows a pattern of bias similar to that seen in the other samples
sequenced on an Illumina platform (see, Supplementary Section 5).

To evaluate the efficacy of each of the bias correction methods considered, we
counted reads overlapping each gene, defining the gene by the union of every
transcript in release 62 of the Ensembl gene annotations, excluding UTRs, and
100 additional bases from both ends of the gene. Counts were then normalized by
dividing by the length of these genes. We then removed any genes with a read
count less than 10, or that did not correspond to

Each method was trained in a manner identical to that used in the analysis of
section 3.1 and 3.2, but without restricting the training data to the first
eight chromosomes.  After adjusting read counts according to the predicted
sequence bias, we computed the Pearson correlation between log read counts and
log TaqMan expression values, which are averaged across three replicates. These
correlations are listed in Table \ref{tab:pcr}.

>>>>

%\title{Correcting for bias in high-throughput sequencing data}
\title{A new approach to bias correction in high-throughput sequencing data}
\author[Jones \textit{et~al}]
{Daniel C. Jones\,$^{1,}$
\hspace{-0.5em}\footnote{to whom correspondence should be addressed}\hspace{0.5em},
Walter L. Ruzzo\,$^{1,2,3}$,
Xinxia Peng$^{4}$
and
Michael G. Katze$^{4}$
}

@@

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated by protocol-specific sequence bias. The exact sources of the
bias are unknown, but may be influenced by PCR amplification, or differing
primer affinities and mixtures, for example.  The result is decreased accuracy
in many applications, such as de novo gene annotation, isoform quantification in
RNA-Seq, and peak calling in ChIP-Seq.

\section{Results:}
We present a new method to measure and correct for these influences using a
simple graphical model. Our model does not rely on existing gene annotations,
making it applicable to any quantitative high throughput sequencing data. We
evaluate our method on several data sets, and by multiple criteria,
demonstrating that it effectively decreases bias and increases uniformity.

@@

In the last few years, RNA-Seq has emerged as a promising alternative to
microarrays in quantifying RNA abundance. But, as microarray technology has
brought with it technical challenges ranging from developing robust
normalization to accounting for cross-hybridization, RNA-Seq presents a new set
of challenges. A significant example is the effect that transcript length has in
biasing quantification, as first explored by \citet{Oshlack2009}, and
subsequently addressed by \cite{Young2010, Gao2011}.

The underlying nucleotide sequence also influences quantification in a manner
that is complex and protocol-specific. In an ideal experiment, the number of
reads mapping to a particular position in the genome is a function of RNA
abundance and should not be additionally dependent on the sequence at that
position. As noted first by \citet{Dohm2008}, this is not the case.  As
illustration, Figure \ref{fig:freqs} plots this nonuniformity in nucleotide
frequencies on five data sets (see Table \ref{tab:datasets}), each using a
different protocol.

@@

Another model, proposed by \citet{Hansen2010}, directly estimates the
distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution, sampled from the ends of reads. The read count
at a given position is then adjusted by the ratio of the foreground and
background heptamer probabilities. Specifying two distributions over heptamers
(i.e., foreground and background distributions) requires \mbox{$2(4^7-1) =
32,766$} parameters, so while no gene annotations are needed to train such a
model, a significant number of reads are required, and a number that increases
exponentially with $k$, if it were desirable to model $k$-mers for $k > 7$.

@@

We begin with a natural model of an RNA-Seq experiment (and one that is often
assumed, whether implicitly or otherwise). The number of reads $x_i$ aligned to
genomic position $i$ is an unbiased estimate of RNA abundance. Furthermore, we
assume reads may be treated as independent and identically distributed samples.
That is, if $N$ reads are generated, and $m_i$ is the event that a generated
read maps to position $i$, then
$$ E[x_i] = N \Pr[m_i] $$

@@

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias. The predicted bias can then be
used to adjust read counts, as in (e). In (d) and (e) we show the results of
this method applied to the $3'$ UTR of Apoa2, using data from
\citet{Mortazavi2008}. In bias coefficients predicted across 10 million positions
of chromosome 1, the log10 bias of 95\% of the positions were between -1.14 and
0.63, suggesting that most adjustments are not large. The $R^2$ measure,
detailed in Section 3.2, gives the relative increase in log-likeliood under a
uniform sampling model, after correcting for bias, with 1.0 indicating a perfect
fit, and the score of 0.38 here indicating a significant increase.
}
\label{fig:overview}
\end{figure}

@@

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity heuristically using the Bayesian
Information Criterion \citep{Schwarz1978a}. Where $m$ is the number of
parameters needed to specify the model, we maximize, 
$ \ell' = 2 \ell - m \log n $.

@@

The run time of the training procedure is further reduced in practice by
imposing the following two restrictions on the structure of the network,
\begin{enumerate}
\item The in-degree (i.e., number of parents) of any node must be less than some
number $p_{\text{max}}$.
\item $|j - i| \le d_{\text{max}}$ for all edges $(i,j)$ and some number $d_{\text{max}}$.
\end{enumerate}
The second of these encodes the assumption that distant nucleotides are
effectively independent. We choose $p_{\text{max}} = 4$ and $d_{\text{max}} =
10$, as reasonable default values.

@@

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing three assumptions we make of an ideal, unbiased experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}), measured
from reads within exons, should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across a single exon should follow, approximately, a Poisson
process.
\item Adjusting for bias in RNA-Seq should increase the agreement between RNA-Seq and
an other methods of quantification.
\end{enumerate}

Evident from Figure \ref{fig:overview}, the assumption of uniform read coverage
often does not hold in typical RNA-Seq data sets. Although the bias corrected
read counts across the exon pictured in this example are visibly more uniform,
we sought a simple, objective tests that could be applied genome-wide. To this
end, we used cross-validation tests (i.e., methods were trained and tested on
disjoint subsets of the same RNA-Seq data sets) of a quantitative measure of the
increase in uniformity of nucleotide frequencies (Kullback-Leibler divergence in
Section 3.1) and increase in uniformity of read coverage (Poisson regression in
Section 3.2). Additionally, we compare RNA-Seq based estimate of gene expression
to qRT-PCR based estimates for the same genes, showing increased correlation
between the two methods after bias correction (Section 3.3).

@@

\begin{table}
\processtable{Data sets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllcc}\toprule
 & & & Read  \\
Experiment & Species & Platform & length \\\midrule
\citet{Wetterbom2010} & Chimpanzee & ABI & 33  \\
Katze (unpublished) & Macaque & ABI & 50  & \\
\citet{Bullard2010} & Human & Illumina & 35 \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
\end{tabular}
}{}
\end{table}

\begin{table}
\processtable{Numbers of reads used to train each method.\label{tab:counts}}
{
\begin{tabular}{lcccc}\toprule
Experiment & Total & 7mer & GLM/MART & BN \\\midrule
Wetterbom & $3.9 \times 10^7$ & $2.3 \times 10^7$ & $5.9 \times
10^6$ & $1 \times 10^5$ \\
Katze & $4.8 \times 10^7$ & $2.0 \times 10^7$ & $3.1 \times 10^6$ & $1 \times
10^5$ \\
Bullard & $1.2 \times 10^7$ & $3.1 \times 10^6$ & $1.8 \times 10^5$ & $1 \times
10^5$\\
Mortazavi & $3.9 \times 10^6$ & $1.9 \times 10^6$ & $3.8 \times 10^5$ & $1
\times 10^5$ \\
Trapnell & $3.4 \times 10^7$ & $1.6 \times 10^7$ & $5.9 \times 10^6$ & $1 \times
10^5$
\end{tabular}
}{Note that in the Trapnell data set (the only paired-end set considered) we
count only the first mate of each read.}
\end{table}

@@

The differences in the methods being tested necessitated training procedures
unique to each. The total number of reads used to train each method is listed in
Table \ref{tab:counts}, and below we describe the procedure used for
each.

@@

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as
$$D_k( f, f' ) = \sum_{x} \left( f_x \log_{2}( f_x / f'_x ) + f'_x \log_{2}( f'_x / f_x) \right)$$
where the sum is over all $k$-mers. This can be thought of as a measure
dissimilarity between two probability distributions. If $f_x$ and $f'_x$ for a
$k$-mer $x$ are approximately equal, their log-ratio will be approximately
zero, leading to a small KL divergence (exactly zero, when the distributions are equal).
Conversely, very different $k$-mer frequencies will result in a larger KL
divergence.

@@

In this comparison, we measure the uniformity of the data, or more precisely
how well the counts conform to a Poisson process.  The assumption of positional
read counts following a Poisson distribution is known to be a poor fit
\citep{Srivastava2010}, but measuring the improvement in the fit derived from
correcting for bias remains a principled and easily interpreted criterion. This
increase in uniformity is illustrated in Figure \ref{fig:overview}.

@@

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a perfect fit, occurring when the model being
evaluated achieves a likelihood of 1. Smaller number indicate an increasingly
worse fit, with $R^2 = 0$ representing no improvement over the null model, and
$R^2 = 0.5$, for example, indicating the model has a log-likelihood equal to
half that of the null model (a large improvement, corresponding to, for example,
the likelihood increasing over 100-fold if the initial log-likelihood was -9.6,
which is the mean per-position log-likehood under the null model). This measure
can also take on values less than 0, indicating that the model has worse fit
than the null model (i.e., when adjusting read counts according the bias
coefficients leads to less uniform read coverage).

We compute $R^2$ for each of the test exons, giving us a sense of the variability
of the effectiveness of each model. The results of this analysis are plotted in
Figure \ref{fig:pois}.  To summarize each model with a single number, we can
examine the median $R^2$ value, as listed in Table \ref{tab:pois}. Our method
shows a highly statistically significant improvement in performance over other
methods in all but one comparison, in which the MART method performs equally.

\begin{figure}
\centerline{\includegraphics[width=0.30\textwidth]{pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we compute McFadden's pseudo-coefficient of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive, and increases as uniformity is
increased, and negative when uniformity is decreased.  Marked with asterisks are
methods over which the BN approach showed a statistically significant improvement
when applied to the same data, according to a one-sided Wilcoxon signed-rank
test. In each of those marked, we observed p-values less than $10^{-23}$.  Boxes
are plotted to mark the 25\%, 50\%, and 75\% quantiles, with whiskers extending
to 1.5 times the inter-quartile range (i.e., the span between the 25\% and 75\%
quantiles), and dots marking more extreme values.
}
    \label{fig:pois}
\end{figure}

\begin{table}
\processtable{
    The median $R^2$ goodness of fit statistic across test exons.
\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & \textbf{0.174} & 0.016 & 0.066 & -0.079 \\
Katze     & \textbf{0.280} & 0.243 & 0.158 & 0.033 \\
Bullard   & \textbf{0.267} & 0.163 & 0.224 & 0.157 \\
Mortazavi & \textbf{0.240} & 0.210 & 0.197 & 0.091 \\
Trapnell  & \textbf{0.289} & \textbf{0.289} & 0.248 & 0.138
    \end{tabular}
}
{
    The $R^2$ statistic measures increased uniformity in read coverage, after
    correcting for bias.  Here the median $R^2$ across the test exons is listed
    for each method and sample. A higher $R^2$ indicates a better fit. The
    highest value in each row is highlighted in bold.
}
\end{table}

@@

We used sequencing data previously published by \citet{Au2010} to evaluate the
effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
made available by the the Microarray Quality Control project \citep{Shi2006}. The
RNA-Seq data shows a pattern of bias similar to that seen in the other samples
sequenced on an Illumina platform (see Supplementary Section 5). This evaluation
does not rely on an assumption that qRT-PCR necissarily more accurate than
RNA-Seq based quantification, only that qRT-PCR is not biased in the same way as
the RNA-Seq data.

To evaluate the efficacy of each of the bias correction methods considered, we
counted reads overlapping each gene, defining the gene by the union of every
transcript in release 62 of the Ensembl gene annotations, excluding UTRs, and
100 additional bases from both ends of the gene. Counts were then normalized by
dividing by the length of these genes. We then removed any genes with a read
count less than 10, or that did not correspond to a unique TaqMan probe.
